---
title: "Introduction to ggplot"
author: "36-290 -- Statistical Research Methodology"
date: "Week 2 Thursday -- Fall 2019"
output: 
  slidy_presentation:
    font_adjustment: -1
---

Preliminaries
===

Let's read in the data frame that we use in the `dplyr` notes set:
```{r}
df = read.csv("http://www.stat.cmu.edu/~pfreeman/GalaxyMass.csv")
```
As a reminder, these data consist of 3456 rows and 10 columns with names
```{r}
names(df)
```

ggplot
===

`ggplot` (actually, and perhaps confusingly, `ggplot2`) is "a system for declaratively creating graphics, based on [The Grammar of Graphics](https://www.amazon.com/Grammar-Graphics-Statistics-Computing/dp/0387245448/ref=as_li_ss_tl?ie=UTF8&qid=1477928463&sr=8-1&keywords=the+grammar+of+graphics&linkCode=sl1&tag=ggplot2-20&linkId=f0130e557161b83fbe97ba0e9175c431). You provide the data [frame], tell `ggplot2` how to map variables to aesthetics, what graphical primitives to use, and it takes care of the details."

Sounds good. Let's dive in:
```{r}
library(ggplot2)    # also loaded as part of the tidyverse
```

ggplot: Basic Structure
===

A very basic call to `ggplot()` has the following structure:
```
ggplot(data=<data frame>,mapping=aes(x=<x axis variable>,...)) + geom_<plot type>(<arguments>)
```
For instance, to plot `z.mode` vs. `mass` (and remember: we plot $y$ vs. $x$):
```{r fig.height=3,fig.width=3,fig.align="center"}
ggplot(data=df,mapping=aes(x=mass,y=z.mode)) + geom_point(color="red")
```
We'll talk more about two-dimensional scatter plots and such in another set of notes. Here, we'll concentrate on univariate (one-dimensional) plots.

ggplot: Bar Chart
===

How many galaxies are in each field?
```{r fig.height=3,fig.width=3,fig.align="center"}
ggplot(data=df,mapping=aes(x=field)) + geom_bar(color="red",fill="blue")
```
(A bar chart is appropriate when the x-axis variable is categorical and the y-axis variable is quantitative.)

ggplot: Histogram
===

What is the distribution of galaxy sizes?
```{r fig.height=3,fig.width=3,fig.align="center"}
ggplot(data=df,mapping=aes(x=size)) + geom_histogram(color="blue",bins=25)
```
(A histogram is appropriate when the single variable in question is quantitative.)

ggplot: Boxplot
===

Boxplots are just a bit trickier. What is the distribution of galaxy concentrations?
```{r fig.height=3,fig.width=3,fig.align="center"}
ggplot(data=df,mapping=aes(x="",y=C)) + geom_boxplot(color="blue",fill="orange")
```
(A boxplot is also for visualizing a quantitative variable.)

ggplot: Faceting
===

Faceting is the act of making multiple plots at once that appear side-by-side as "facets". Faceting is something you might want to do when, e.g., you have a factor variable. Here, we show histograms of the concentration variable `C` broken up by galaxy field.
```{r fig.height=4,fig.width=4,fig.align="center"}
ggplot(data=df,mapping=aes(x=C)) + geom_histogram(color="blue",fill="yellow",bins=25) +
  facet_wrap(~field,scales='free_x')
```

ggplot: Gather
===

`gather()` is a function (from the `tidyr` package) that takes a data frame and realigns it. It is best illustrated via a simple example. Let's say we have the following data frame, which we'll call `df`:
```
> df
     x   y
1  0.5 0.7
2  1.2 1.9
```
If we "gather" these data, we get the following:
```
> library(tidyr)
...
> gather(df)
      key  value
1       x    0.5
2       x    1.2
3       y    0.7
4       y    1.9
```
Combining `gather()` with faceting allows one to, e.g., visualize multiple variables at once.

ggplot: Gather (+ dplyr)
===

```{r fig.height=3,fig.width=4,fig.align="center"}
suppressMessages(library(tidyr))
suppressMessages(library(magrittr))
suppressMessages(library(dplyr))

df.new = df %>% select(.,z.mode,mass) %>% gather(.)
ggplot(data=df.new,mapping=aes(x=value)) + geom_histogram(color="blue",fill="yellow",bins=25) +
  facet_wrap(~key,scales='free_x')
```

ggplot: Scatter Plot
===

Let's get a sense of the data by plotting `mass` vs. `C` in a scatter plot:
```{r fig.height=5,fig.width=5,fig.align="center"}
ggplot(data=df,mapping=aes(x=C,y=mass)) + geom_point(color="red",size=0.5)
```

ggplot: Scatter Plot with Density
===

By using `geom_density_2d()` we can overlay contours that indicate the density of points:
```{r fig.height=5,fig.width=5,fig.align="center"}
ggplot(data=df,mapping=aes(x=C,y=mass)) + geom_density_2d() + geom_point(color="red",size=0.5)
```

ggplot: Scatter Plot with Bins
===

By using `geom_bin2d` we can overlay bins whose color indicates the density of points:
```{r fig.height=5,fig.width=5,fig.align="center"}
ggplot(data=df,mapping=aes(x=C,y=mass)) + geom_bin2d()
```

ggplot: Scatter Plot with Hexagonal Bins
===

What about hexagonal bins?
```{r fig.height=5,fig.width=5,fig.align="center"}
ggplot(data=df,mapping=aes(x=C,y=mass)) + geom_hex() + geom_point(color="red",size=0.25)
```

ggplot: freqpoly
===

`geom_freqpoly()` is a means by which to overlay empirical distributions of data in one plot pane. For instance, what is the distribution of `mass` as a function of `field`?
```{r fig.height=5,fig.width=5,fig.align="center"}
ggplot(data=df,mapping=aes(x=mass)) + geom_freqpoly(mapping=aes(color=field))
```

ggplot: Overlaid Histograms
===

```{r fig.height=5,fig.width=5,fig.align="center"}
ggplot(data=df,mapping=aes(x=C,fill=field,stat(density))) + 
  geom_histogram(alpha=0.4,position="Identity")
```
In this case: not particularly helpful. `freqpoly` is better here.

ggplot: Side-By-Side Boxplots
===

```{r fig.height=5,fig.width=5,fig.align="center"}
ggplot(data=df,mapping=aes(x=field,y=mass)) + geom_boxplot(fill="FireBrick2") 
```
