---
title: "Decision Trees: Examples"
author: "Peter Freeman"
date: "Summer 2019"
output: 
  slidy_presentation:
    font_adjustment: -1
---

Regression Example
===

The dataset loaded below contains brightness, distance, and mass information for 10,000 simulated galaxies. In our regression tree example, we will try to relate the measurements of brightness in six different bandpasses (`u`,...,`y`) to the measure of distance (`redshift`).
```{r}
df = read.csv("https://raw.githubusercontent.com/pefreeman/PSU_2019/master/PhotoZ.csv")
names(df)

set.seed(808)
s = sample(nrow(df),0.7*nrow(df))
pred.train = df[s,1:6]
pred.test  = df[-s,1:6]
resp.train = df[s,7]
resp.test  = df[-s,7]
```

Regression Example
===

There is more than one package that one can use to learn a regression tree; here we use the `rpart` package.
```{r fig.height=6,fig.width=6,fig.align="center"}
if ( require(rpart) == FALSE ) {
  install.packages("rpart",repos="https://cloud.r-project.org")
  library(rpart)
}
if ( require(rpart.plot) == FALSE ) {
  install.packages("rpart.plot",repos="https://cloud.r-project.org")
  library(rpart.plot)
}

out.rpart = rpart(resp.train~.,data=pred.train)
resp.pred = predict(out.rpart,newdata=pred.test)
mean((resp.pred-resp.test)^2)
```

Regression Example
===

```{r fig.height=6,fig.width=6,fig.align="center"}
library(ggplot2)

ggplot(data=data.frame(resp.test,resp.pred),mapping=aes(x=resp.test,y=resp.pred)) + geom_point() +
  xlim(0,2) + ylim(0,2) + geom_abline(intercept=0,slope=1,color="red")
```

Regression Example
===

```{r fig.height=6,fig.width=6,fig.align="center"}
rpart.plot(out.rpart)
```

Classification Example
===

The dataset that we read in below contains magnitude and redshift data for 500 quasars and 500 stars. The idea is to learn a classifier that can discriminate between quasars and stars with a low misclassification rate.
```{r}
df = read.csv("https://raw.githubusercontent.com/pefreeman/PSU_2019/master/StarQuasar.csv")
dim(df)
names(df)
set.seed(202)
s = sample(nrow(df),0.7*nrow(df))
pred.train = df[s,1:5]                                  # don't include redshift or redshift error!
pred.test  = df[-s,1:5]
resp.train = df[s,8]
resp.test  = df[-s,8]
```

Classification Example
===

```{r}
out.rpart = rpart(resp.train~.,data=pred.train)
resp.pred = predict(out.rpart,newdata=pred.test,type="class")
(t = table(resp.pred,resp.test))
(t[1,2]+t[2,1])/sum(t)
```

Classification Example
===

```{r fig.height=6,fig.width=6,fig.align="center"}
rpart.plot(out.rpart)
```

