---
title: "Beyond ggplot: Extra Visualization Resources"
author: "36-290 -- Statistical Research Methodology"
date: "Week 3 Tuesday -- Fall 2020"
output: 
  slidy_presentation:
    font_adjustment: -1
---

Preliminaries
===

Let's read in the data frame that we use in the `dplyr` and `ggplot` notes sets:
```{r}
df = read.csv("http://www.stat.cmu.edu/~pfreeman/GalaxyMass.csv")
```
As a reminder, these data consist of 3456 rows and 10 columns with names
```{r}
names(df)
```

Covariance and Correlation
===

Covariance is a measure of the *linear* dependence between two variables. (To be "uncorrelated" is not the same as to be "independent"...the latter means there is no dependence, linear or otherwise, between two variables.) Correlation is a "normalized" form of covariance, that ranges from -1 (one variable linearly decreases absolutely in value while the other increases in value) through 0 (no linear dependence) to 1 (one variable linear increases absolutely while the other increases).

A good package for visualizing correlation is `corrplot`:
```{r}
if ( require(corrplot) == FALSE ) {
  install.packages("corrplot",repos="https://cloud.r-project.org")
  library(corrplot)
}
```

corrplot
===

Here we generate a correlation plot for the galaxy morphological variables  `Gini`, `M20`, `C`, and `A`:
```{r fig.height=5,fig.width=5,fig.align="center"}
library(tidyverse)
df %>% select(.,Gini,M20,C,A) %>% cor(.) %>% corrplot(.,method="ellipse")
```

GGally: Pairs Plots
===

We step outside canonical `ggplot` plotting for now to bring you a pairs plot from the `GGally` package:
```{r fig.height=7,fig.width=7,fig.align="center"}
if ( require(GGally) == FALSE ) {
  install.packages("GGally",repos="https://cloud.r-project.org")
  library(GGally)
}
df %>% select(.,Gini,M20,C,A) %>% ggpairs(.,progress=FALSE,lower=list(combo=wrap("facethist", binwidth=0.8)),color="blue")
```

3D Scatter Plot
===

We now look at methods for visualizing data in three and more dimensions. The first is via the `scatterplot3d` package:
```{r fig.height=5,fig.width=5,fig.align="center"}
if ( require(scatterplot3d) == FALSE ) {
  install.packages("scatterplot3d",repos="https://cloud.r-project.org")
  library(scatterplot3d)
}
scatterplot3d(x=df$Gini,y=df$M20,z=df$C,pch=19,color="blue",angle=45,cex.symbols=0.5)
```

rgl: R Graphics Library
===

Wouldn't it be nice if we could interactively control the 3D scatterplot? Hmmmm...
```{r}
if ( require(rgl) == FALSE ) {
  install.packages("rgl",repos="https://cloud.r-project.org")
  library(rgl)
}
plot3d(x=df$Gini,y=df$M20,z=df$C,pch=19)
f = spin3d(axis=c(0,0,1),rpm=5)
play3d(f,duration=10)
```

Parallel Coordinates
===

The `parcoord()` function in the `MASS` package is one mechanism through which we can attempt to visualize more than three variables at once (although we only visualize three below). Each line in a parallel coordinates plot represents a single object. For instance, in the plot below, we find that there is a dense set of lines that are colored red: these are layups and slam dunks, shots that cluster tightly in all three dimensions.
```{r}
library(MASS)  # this should be in your base distribution
z.color = round(64*df$Gini)
palette(rainbow(64))
parcoord(df[,c("Gini","M20","C","A")],col=z.color,lwd=0.6)
```

